---
title: "Module 3 - Assignment"
output: 
  learnr::tutorial:
    progressive: true
    allow_skip: true
    df_print: paged
runtime: shiny_prerendered
description: >
  Assignment for Data Management in R module
---

```{r setup, include=FALSE, message=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(learnr)
library(tidyverse)
library(randomNames)

set.seed(1)

farmer_data <- data.frame(
  id  = sample(1:100, size = 100),
  village_id = sample(1:8, size = 100, replace = TRUE),
  name = c(randomNames(50, gender = 0, which.names = "first"), randomNames(50, gender = 1, which.names = "first")),
  sex = c(sample(c("1", "M", "m", "Male"), 50, replace = TRUE), sample(c("2", "F", "f", "Female"), 50, replace = TRUE)),
  age = c(sample(seq(from = 18, to = 65, by = 1), 99, replace = TRUE), 180),
  hh_size = sample(c(1:10, -9), size = 100, replace = TRUE),
  income = sample(c(seq(from = 100, to = 900, by = 60), 0, -9), size = 100, replace = TRUE),
  costs = sample(c(seq(from = 50, to = 500, by = 20), -9), size = 100, replace = TRUE)
)

xx <- sample(1:100, 5)

for(i in 1:5){
  farmer_data$name[xx[i]] <- paste0(" ", farmer_data$name[xx[i]])
}

yy <- 1:100

yy <- yy[-xx]

yy <- sample(1:100, 5)

for(i in 1:5){
  farmer_data$name[yy[i]] <- paste0(farmer_data$name[yy[i]], " ")
}

plot_data <- data.frame(
  id = 1:300,
  farmer_id = sample(1:100, 300, replace = TRUE),
  area = sample(seq(from = 0.4, to = 8, by = 0.4), 300, replace = TRUE),
  pesticides = sample(0:1, 300, replace = TRUE),
  fertilisers = sample(0:1, 300, replace = TRUE)
)

plot_data <- plot_data%>%
  group_by(farmer_id)%>%
  mutate(plot_num = row_number())

plot_data <- pivot_wider(plot_data,
            id_cols = farmer_id,
            names_from = plot_num,
            values_from = c(id, area:fertilisers),
            names_sep = "_")

village_data <- data.frame(
  id = rep(c(1:8),3),
  name = rep(c("A","B","C","D","E","F","G","H"),3),
  information = c(rep("population", 8), rep("ngo_supported",8), rep("weight",8))
)

for(i in 1:nrow(village_data)){
  village_data$value[i] <- ifelse(
    village_data$information[i] == "population", sample(1500:10000,size = 1),
    ifelse(village_data$information[i] == "ngo_supported", sample(c("Yes", "No")), 
           sample(seq(0.02,1.98, by = 0.02)))
    )
}

farmer_data_additional <- data.frame(
  id  = 101:120,
  village_id = sample(1:8, size = 20, replace = TRUE),
  name = c(randomNames(10, gender = 0, which.names = "first"), randomNames(10, gender = 1, which.names = "first")),
  sex = c(rep("Male", 10), rep("Female", 10)),
  age = c(sample(seq(from = 18, to = 65, by = 1), 20, replace = TRUE)),
  hh_size = sample(1:10, size = 20, replace = TRUE),
  income = sample(c(seq(from = 100, to = 900, by = 60), 0), size = 20, replace = TRUE),
  costs = sample(seq(from = 50, to = 500, by = 20), size = 20, replace = TRUE)
)

farmer_data_additional$gross_income = farmer_data_additional$income - farmer_data_additional$costs

plot_data_additional <- data.frame(
  farmer_id = sample(101:120, size = 45, replace = TRUE),
  id = 301:345,
    area = sample(seq(from = 0.4, to = 8, by = 0.4), 45, replace = TRUE),
  pesticides = sample(0:1, 45, replace = TRUE),
  fertilisers = sample(0:1, 45, replace = TRUE)
)

plot_data_additional <- plot_data_additional%>%
  group_by(farmer_id)%>%
  mutate(plot_num = as.character(row_number()))%>%
  arrange(farmer_id, plot_num)

plot_data_long <-plot_data%>%
  pivot_longer(
    cols = id_1:fertilisers_8,
    names_to = c(".value", "plot_num"),
    names_sep = "_",
    values_drop_na = TRUE
  )

village_data_wide <- village_data%>%
  pivot_wider(
    names_from = information,
    values_from = value
  )

farmer_data_clean <- farmer_data%>%
  mutate(name = trimws(name, which = "both"),
         hh_size = na_if(hh_size, -9),
         income = na_if(income, -9),
         costs = na_if(costs, -9),
         sex = case_when(
    sex %in% c("Male", "m", "M", "1") ~ "Male",
    sex %in% c("Female", "f", "F", "2") ~ "Female"
  ),
  age = ifelse(age == 180, 18, age),
  gross_income = income-costs)%>%
  arrange(id)

plot_data_summary <- plot_data_full%>%
  group_by(farmer_id)%>%
  summarise(total_area = sum(area, na.rm = TRUE),
            nplots = n())

farmer_data_full <- bind_rows(farmer_data_clean, farmer_data_additional)

plot_data_full <- bind_rows(plot_data_long, plot_data_additional)


```

## Part 1 - Data Cleaning

### Dataset

To start with you will work on the following data set which contains a number of variables for a list of 100 farmers.

```{r, exercise = FALSE}
farmer_data
```

### Exercise 1a - Trim Whitespace

Let's start by checking for any whitespace, i suspect there may be some in our name column.

Let's perform a check by using the `unique` function to look at all the unique values for name

```{r, exercise = FALSE}
unique(farmer_data$name)
```

Seems our suspicions were correct, we have around 10 values for name that have either a leading or trailing space in them.

Exercise 1a: Using `mutate` and `trimws`, remove the whitespace present in the name column from both the beginning and the ending of the variable entries

**REMEMBER** You can inspect the results of your changes at any time by writing the name of the data frame into the code chunk at any time

```{r ex1a, exercise = TRUE}
farmer_data <- farmer_data %>%
  mutate()

unique(farmer_data$name)
```

```{r ex1a-solution}
farmer_data <- farmer_data %>%
  mutate(name = trimws(name, which = "both"))

unique(farmer_data$name)
```

### Exercise 1b - Missing Data

Recall that in R we need to be very explicit when it comes to defining data. It **MUST** be coded as `NA`

In our dataset, we have unfortunately not told R explicitly what our missing data codes were when we imported the data. Therefore, there are a few variables which have some values equalling -9. However this was a missing data code, but R will see this value as literally -9.

Let's check first which variables this issue applies to.

Exercise 1bi: Use `summary` to check which continuous variables in the dataset are still using the missing data code `-9` rather than being explicitly `NA`. Note that -9 will be the minimum value in each of these cases.

```{r ex1bi, exercise = TRUE}


```

```{r ex1bi-solution}
summary(farmer_data)

farmer_data

# Should have found that at least 1 value in each of hh_size, income and costs were equal to -9
```

You should have identified a few variables unfortunately using this code.

Exercise 1bii: Using `mutate` and `na_if`, correct the variables that are still using `-9` so that these are instead coded as `NA`

```{r ex1bii, exercise = TRUE}
farmer_data <- farmer_data %>%
  mutate()

summary(farmer_data)
```

```{r ex1bii-solution}
farmer_data <- farmer_data %>%
  mutate(hh_size = na_if(hh_size, -9),
         income = na_if(income, -9),
         costs = na_if(costs, -9))

summary(farmer_data)

# OR

farmer_data <- farmer_data %>%
  mutate_all(na_if, -9)

summary(farmer_data)

#This would only be useful if -9 will ALWAYS be a missing code and never a true value of a variable

#Recommend looking into mutate_at or mutate_if to see how you can make the same edit across multiple but not all columns
```

### Exercise 1c - Correcting labels

Now something else looked off about our sex variable. Let's use `unique` to have a look at all of the unique values of sex

```{r, exercise = FALSE}
unique(farmer_data$sex)
```

Hmm, seems we have multiple values to mean the same thing unfortunately. Male and Female have both been coded four different ways each.

Exercise 1c: Using `mutate` and `case_when`, standardise the sex variable such that

Male, m, M & "1" are all recoded to equal "Male"

Female, f, F & "2" are all recoded to equal "Female"

**Hint** Think back to how we used the `%in%` operator

```{r ex1c, exercise = TRUE}
farmer_data <- farmer_data %>%
  mutate()
```

```{r ex1c-solution}
farmer_data <- farmer_data %>%
  mutate(sex = case_when(
    sex %in% c("Male", "m", "M", "1") ~ "Male",
    sex %in% c("Female", "f", "F", "2") ~ "Female"
  ))

table(farmer_data$sex)
```

### Exercise 1d - Correcting outlier values

Let's use a histogram to inspect one of our continuous variables for possible data entry errors

```{r}
hist(farmer_data$age)
```

Looks like we have one implausibly high outlying value for age.

Upon contacting the enumerator who collected data for this farmer, you find the true value was meant to be 18 years It was just inputted wrongly.

Exercise 1d: Using any method demonstrated in the workbook, correct this implausibly high value of 180 years old to the correct value of 18 years old.

```{r ex1d, exercise = TRUE}

```

Remember you can use the histogram to again plot the value and check your changes

```{r ex1db, exercise = TRUE}
hist(farmer_data$age)
```


```{r ex1d-solution}
farmer_data$age[farmer_data$age == 180] <- 19

#OR

farmer_data <- farmer_data%>%
  mutate(age = ifelse(age == 180, 18, age))
```

### Exercise 1e - Creating a new variable

As part of the data analysis plan you notice that part of the analysis will include an evaluation of the gross income of farmers, a calculation of income minus costs. However, you find this was not calculated directly as part of the data collection.

Exercise 1e: Using `mutate` create a new variable called `gross_income` which is the income minus the costs.

```{r ex1e, exercise = TRUE}

```

```{r ex1e-solution}
farmer_data <- farmer_data%>%
  mutate(gross_income = income-costs)

farmer_data$gross_income
```

### Exercise 1f - Sorting data

Finally, you may have noticed that there is currently no particular order to our data. While not mandatory, sorting data can be useful for presentation, readability and help users easily find certain pieces of information.

We have a natural sorting variable in our unique identifier "id"

Exercise 1f: Using `arrange`, sort the data using the id variable so that the rows are in order from 1 to 100 by id.

```{r ex1f, exercise = TRUE}

```

```{r ex1f-solution}
farmer_data <- farmer_data%>%
  arrange(id)
```

```{r, echo = FALSE}
farmer_data <- farmer_data%>%
  mutate(name = trimws(name, which = "both"),
         hh_size = na_if(hh_size, -9),
         income = na_if(income, -9),
         costs = na_if(costs, -9),
         sex = case_when(
    sex %in% c("Male", "m", "M", "1") ~ "Male",
    sex %in% c("Female", "f", "F", "2") ~ "Female"
  ),
  age = ifelse(age == 180, 18, age),
  gross_income = income-costs)%>%
  arrange(id)
```

## Part 2 - Data Reshaping

### Datasets

Before we test your skills in pivoting data, let’s look at some additional data we have for this project. It is at the level of the plot while our previous dataset was at the level of the farmer. 

Each farmer can have up to 8 plots, they are labelled id_1, id_2 and so on. You will notice that many of the values in these columns are NA as the farmers do not all have this many plots.`

```{r}
plot_data
```

Secondly, you have been provided by an outside source with some data about the villages that have been used in this project. However, it has unfortunately been passed along in long format with individual rows for the villages' population, sampling weights and whether or not they received support from an NGO.

```{r}
village_data
```

### Exercise 2a - Pivot longer

Firstly, let's get our plot data into a more useful shape. When conducting analysis it will be much simpler to keep comparable information into the same column. i.e keep all values for the area of plot in just the single column.

Exercise 2a: Using `pivot_longer` reshape the wide plot data into a new long data set, save this output as `plot_data_long` and call the new variable, that will list the plot number within a farmer, "plot_num". 

Please include the `values_drop_na = TRUE` argument into your answer. This will stop unnecessary rows being created for where all variables will be `NA`. In other words, all farmers will only have as many rows as they will plots.

**Hint**: If you are stuck at all, look back at the session 2 workbook and how we utilised the special string of `".value"`.

**HINT** For the following steps you can use `dim()` to check the number of rows and columns in your data to make sure you have the right amount.

```{r ex2a, exercise = TRUE}
plot_data_long <-plot_data%>%
  pivot_longer(
    values_drop_na = TRUE
  )
```

```{r ex2a-solution}
plot_data_long <-plot_data%>%
  pivot_longer(
    cols = id_1:fertilisers_8,
    names_to = c(".value", "plot_num"),
    names_sep = "_",
    values_drop_na = TRUE
  )

plot_data_long

dim(plot_data_long)
```


### Exercise 2b - Pivot wider

Next let's do the opposite to that village data.

Exercise 2b: Using `pivot_wider`, transform the data into wide format such that you only have one row per village. Save this to a new object titled `village_data_wide`



```{r ex2b, exercise = TRUE}
village_data_wide <- village_data%>%
  pivot_wider()
```

```{r ex2b-solution}
village_data_wide <- village_data%>%
  pivot_wider(
    names_from = information,
    values_from = value
  )

village_data_wide

dim(village_data_wide)
```

## Part 3 - Data Merging

### Datasets

For the first exercise of this final section, we are going to bind our previous data sets with some additional observations for our farmers and plots. 20 new farmers with 45 plots between them.

```{r}
farmer_data_additional
```

```{r}
plot_data_additional
```

### Exercise 3a - Binding

As the datasets for our additional data sets which use the same formats that we have developed from our tidied up data.  We can move straight to appending them onto our existing data frames.

Exercise 3a: Using `bind_rows`, append on the new observations to farmer_data and plot_data_long.

**NOTE**: The cleaned version of the farmer data has been saved to a new object called `farmer_data_clean`. Please use this dataset instead of `farmer_data`

```{r ex3a, exercise = TRUE}
farmer_data_full <- 
  
plot_data_full

dim(plot_data_full)

dim(farmer_data_full)
```

```{r ex3a-solution}
farmer_data_full <- bind_rows(farmer_data_clean, farmer_data_additional)

plot_data_full <- bind_rows(plot_data_long, plot_data_additional)

dim(plot_data_full)

dim(farmer_data_full)
```

### Exercise 3b - Summarising

Now we will move onto first summarising some data from a lower level to then then join it up with data at a higher level.

So as part of our data analysis plan we need a total farm area which is the sum of the individual plot areas for each individual farmer. We then want to merge this with our farmer data set.

Exercise 3b: Using `group_by` and `summarise`, create a new data set which is a summary of the plot level data. Call this `plot_data_summary`. Create summary variables for the total area of a farmer's plots using `sum` and then also create a variable which counts the number of plots a farmer owns.

```{r ex3b, exercise = TRUE}
plot_data_summary <- plot_data_full%>%
```

```{r ex3b-solution}
plot_data_summary <- plot_data_full%>%
  group_by(farmer_id)%>%
  summarise(total_area = sum(area, na.rm = TRUE),
            nplots = n())

plot_data_summary
```

### Exercise 3c - Joining data

Join the plot_data_summary data set onto the farmer_data_full dataset. Now we actually have a few farmers who we have no plot data available. We have made the decision for analysis that we only want farmers who own at least one plot.

Exercise 3c: Using a type of join that will only keep the farmers who have plot data that they can be matched to, join plot_data_summary onto farmer_data_full. 

**HINT** Remember that you can use `c()` to match variables where they have different names in the two datasets

```{r ex3c, exercise = TRUE}
farmer_data_full <- 
```

```{r ex3c-solution}
farmer_data_full <- inner_join(
  farmer_data_full, plot_data_summary, by = c("id" = "farmer_id")
)
```

Thank you for completing this assignment.

Please now go to the Module 3 quiz where you will be asked a couple questions about the assignment. This quiz will be used to mark completion of the module. Once you have completed the quiz, you can expect to receive your certificate within a few working days.
